---
output: 
  html_document:
    theme: default
    highlight: haddock
    df_print: tibble
---

```{r, echo = FALSE, message = FALSE}

library(dplyr)
library(tidyr)
library(ggplot2)

##############################
######## LOADING DATA ########
##############################

player_value <- read.csv("output//player_value.csv", stringsAsFactors = FALSE)


####################################
######## DATA VISUALIZATION ########
####################################





```

# Player Valuation

### Introduction

Perhaps not surprisingly, I am huge fan of the [sabermetric revolution](https://en.wikipedia.org/wiki/Sabermetrics) within baseball (shocking, I know). And, after figuring out the weekly totals needed to ensure a playoff spot, I decided to see if I could devise a method for evaluating a player's performance. 


### Project Overview

The goal of this project to devise a method of evaluating the performance of a player, such that the resulting metric should be:

* **Intuitive** - Wins (i.e. category points) are the currency of a head-to-head categories league, so the metric should attempt to approximate the number of additional wins a team would gain by having a player on a roster. 
* **Relative to some baseline performance** - In fantasy baseball leagues, team owners can pick up players from the free agency to replace players on their roster. As a result, the metric should be relative to a baseline performance, similar to [Wins Above Replacement (WAR)](https://en.wikipedia.org/wiki/Wins_Above_Replacement).  
* **Based on weekly performance** - The relationship between the weekly totals and the probability of winning a category in a given week exhibits a strong S-shaped curve. As such, players with streaky week-to-week performance won’t win you as many point as a player who puts up similar number, but is more consistent. So the metric should be based on weekly performance, rather than season totals.
* **Based on an average, balance team** - The true value of a player depends on the composition of a team. Paul Goldschmidt is going to add a lot of value to a team that is already stacked with power hitters; nor would he win you a lot of points if your team consisted of only slap hitting, speedsters. To get around this complication, we will assume the team composition has a 50% chance of winning each category when considering a player's contribution 


### Methods




``` {r, eval = FALSE, attr.source = ".numberLines"}

## getting raw change in win probability (i.e. Wins)

totals_batters %>%
  mutate(prob_50   = -coef_Int/coef_Value,
         new_total = case_when(Category == "AVG" ~ (prob_50*avg_AB + H)/(avg_AB + AB), 
                               TRUE ~ prob_50+Total),
         new_prob  = 1/(1+exp(-(coef_Value*new_total + coef_Int))),
         Wins      = new_prob - 0.5
         ) -> totals_batters

totals_pitchers %>%
  mutate(prob_50   = -coef_Int/coef_Value,
         new_total = case_when(Category == "ERA"  ~ (prob_50/9*avg_IP + ER)/(avg_IP + IP)*9, 
                               Category == "WHIP" ~ (prob_50*avg_IP + BR)/(avg_IP + IP),
                               TRUE ~ prob_50+Total),
         new_prob  = 1/(1+exp(-(coef_Value*new_total + coef_Int))),
         Wins      = new_prob - 0.5
         ) -> totals_pitchers

```


#### Replacement Level Performance

After calculating players' raw performance, we need to

In the code below, we calculate 

In a ten team league, there are 25o rosters spots, not counting DL slots, 

``` {r, eval = FALSE, attr.source = ".numberLines"}

## Getting average weekly wins per player

player_value %>%
  filter(Games > 0) %>%
  group_by(Player_ID,
           Player,
           Role,
           Week_ID) %>%
  summarise(Wins = sum(Wins)) %>%
  group_by(Player_ID,
           Player,
           Role) %>%
  summarise(avg_Wins = mean(Wins)) %>%
  ungroup() -> temp_replacements

## Getting list of replacement level players (top 25 P/B outside of the top 250 players)

temp_replacements %>%
  arrange(desc(avg_Wins)) %>%
  filter(row_number() > 250) %>%
  group_by(Role) %>%
  top_n(n  = 25,
        wt = avg_Wins) -> temp_replacements

```


test teeest etset

``` {r, eval = FALSE, attr.source = ".numberLines"}

## Getting average weekly Wins for each category of replacement level players

player_value %>%
  filter(Player_ID %in% temp_replacements$Player_ID,
         Games > 0) %>%
  group_by(Match_Type,
           Role,
           Category) %>%
  summarise(r_Wins = mean(Wins)) -> temp_replacements

## Substracting replacement level performance from raw Wins

player_value %>%
  left_join(y  = temp_replacements,
            by = c("Match_Type", "Role", "Category")) %>%
  mutate(WAR = ifelse(Games == 0, 0, Wins - r_Wins)) %>%
  select(-r_Wins) -> player_value

```


### Results


### Final Remarks



<br>
<br>