---
output: 
  html_document:
    theme: default
    highlight: haddock
    df_print: tibble
---

``` {r message = FALSE, echo = FALSE}

library(dplyr)
library(tidyr)
library(tibble)
library(stringr)
library(kableExtra)
library(knitr)


##############################
######## LOADING DATA ########
##############################

setwd(dirname(getwd()))

## Clean Data

Draft_Results   <- read.csv("data//Clean Data//Draft_Results.csv", stringsAsFactors = FALSE, nrows = 1000)
Matchups        <- read.csv("data//Clean Data//Matchups.csv", stringsAsFactors = FALSE, nrows = 1000)
Weekly_Totals   <- read.csv("data//Clean Data//Weekly_Totals.csv", stringsAsFactors = FALSE, nrows = 1000)
Standings       <- read.csv("data//Clean Data//Standings.csv", stringsAsFactors = FALSE, nrows = 1000)
Week_19_Rosters <- read.csv("data//Clean Data//Week_19_Rosters.csv", stringsAsFactors = FALSE, nrows = 1000)
Final_Rosters   <- read.csv("data//Clean Data//Final_Rosters.csv", stringsAsFactors = FALSE, nrows = 1000)
  
```

# Data & Web Scrapers

### Introduction

I’ve been playing in a fantasy baseball league with some college friends for several years and have generally done pretty well. But in 2018, rather than doing my normal draft prep and research, I decided to take a deeper dive and analyze some fantasy baseball data that I scraped from ESPN.


### Web Scrapers

For this project, I scraped data from ESPN and Baseball Reference. I'm not going to walk through the code (there is a lot of it), but scraping scripts can be found in my [github repo](https://github.com/jason-hanser/fantasy-baseball/tree/main/1%20-%20Scraping%20and%20Cleaning), if you are interested. Because of the format of my league,  I only scraped data from standard, ten team, head-to-head, categories-based leagues. 

NOTE: The scrapers were built in early-2018 and may no longer function properly. 

### Data

Below is a summary of the raw and cleaned data that I collected. All of the data, raw and clean, can be viewed and downloaded from [dropbox](https://www.dropbox.com/home/Public%20Data/Fantasy%20Baseball%20Data). And, within my [github repo](https://github.com/jason-hanser/fantasy-baseball/tree/main/1%20-%20Scraping%20and%20Cleaning), you can find the script I used to clean the data. 

#### Raw Data

The nine tables below are data scraped from the ESPN:

* **Leagues** - A table containing public, ESPN fantasy baseball leagues from the 2017 season, including details about the size and format of each league.
* **League Index** - A table containing the competitiveness of each league (i.e. league index). League index is a uniformly distributed variable, presumably a percentile rank. 
* **League Winners** - A table containing the name of team within each league that won the  championship. 
* **Matchups** - A table containing the weekly totals for each match-up of each team as well as the name of the opposing team. 
* **Matchup Dates** - A small table containing the start and end dates of each match-up. 
* **AB_IP** - A table containing the number of at-bats and innings pitched during a match-up for each team. 
* **Draft Results** - A table containing the draft results for each league, including the draft position of players taken by each team.
* **Final Rosters** - A table containing the roster of each team at the end of the season.  
* **Week 19 Rosters** - A table contain the roster of each team at the end of Week 19. Teams often cannibalize their rosters at the end of the season in a last ditch effort to win, so the final roster is not always reflective of their final roster. 

The two tables below are data scraped from Baseball Reference:

* **Pitcher Game Logs** - Game logs of all MLB pitchers for the 2017 season,
* **Batter Game Logs ** - Game logs of all MLB batters for the 2017 season.

And, finally:

* **Player IDs** - A table containing all of the players within the ESPN and Baseball Reference universes for 2017. This file is used a cross-walk in order to join data from the different sources.


#### Clean Data

The tables below show the first 1,000 rows of each cleaned data file. 

**Draft Results**

<br>
``` {r, echo = FALSE}

kable(x = Draft_Results,
      row.names = FALSE,
      format = "html") %>%
  kable_styling() %>%
  scroll_box(height  = "200px")

```
<br>

**Macth-Ups**

<br>
``` {r, echo = FALSE}

kable(x = Matchups,
      row.names = FALSE,
      format = "html") %>%
  kable_styling() %>%
  scroll_box(height  = "200px")

```
<br>

**Standings**

Contains the weekly standings for each league based on the match-up data. 

<br>
``` {r, echo = FALSE}

kable(x = Standings,
      row.names = FALSE,
      format = "html") %>%
  kable_styling() %>%
  scroll_box(height  = "200px")

```
<br>

**Week 19 Rosters**

<br>
``` {r, echo = FALSE}

kable(x = Week_19_Rosters,
      row.names = FALSE,
      format = "html") %>%
  kable_styling() %>%
  scroll_box(height  = "200px")

```
<br>

**Final Rosters**

<br>
``` {r, echo = FALSE}

kable(x = Final_Rosters,
      row.names = FALSE,
      format = "html") %>%
  kable_styling() %>%
  scroll_box(height  = "200px")

```
<br>

**Weekly Totals** 

Contains the weekly totals of each player, based on Baseball Reference data.

<br>
``` {r, echo = FALSE}

kable(x = Weekly_Totals,
      row.names = FALSE,
      format = "html") %>%
  kable_styling() %>%
  scroll_box(height  = "200px")

```
<br>

### Final Remarks

This was one of the first major web scraping projects I ever undertook. In a few cases, the web scraper returned incomplete or improperly formatted data. Rather than attempting to re-scrape the data from these leagues, I removed those leagues from the dataset. As such, the final dataset contains only those leagues for which I was able to accurately scrape all data, some five thousand plus leagues.

<br>
<br>